Global patterns of forest autotrophic carbon fluxes

### 2020-06-05 manuscript/manuscript draft.Rmd
finish HM comments on intro

Carbon cycling in forests worldwide provides the energetic basis for sustaining the majority of Earth's terrestrial biodiversity and many human populations ([MEA 2005](http://www.millenniumassessment.org/documents/document.354.aspx.pdf); **REFS**), while strongly influencing atmospheric CO~2~ and climate [@bonan_forests_2008]. Forests' autotrophic carbon fluxes (FACF)--that is, carbon fixation, allocation, and metabolism by trees and other primary producers--sets the energy ultimately available to heterotrophic organisms (including microbes), in turn influencing their abundance (e.g., [Zak et al. 1994](http://esajournals.onlinelibrary.wiley.com/doi/abs/10.2307/1940888); [Niedziałkowska et al. 2010](https://doi.org/10.2980/17-1-3310)) and possibly diversity [@waide_relationship_1999; Chu et al. 2019 (DOI: 10.1111/ele.13175)]. FACF influences all organic matter stocks in forest ecosystems and is linked to cycling of energy, water, and nutrients.

@@

FACF decrease with latitude, but it remains unclear whether and how the shape of this relationship varies between fluxes. FACF are lowest in the boreal regions, and increase into the temperate regions [@luyssaert_co_2007; @huston_global_2009; @beer_terrestrial_2010; @piao_forest_2010; @jung_global_2011]. However, evidence is inconclusive on whether primary productivity continues to increase into the tropics. Evidence for this is further complicated by the fact that different studies use different measures of productivity to explore these relationships. For example, modelling of global terrestrial ecosystem gross primary productivity ($GPP$) through upscaling and calibration of eddy flux measurements indicates that GPP peaks in tropical forests [@beer_terrestrial_2010; @jung_global_2011; @badgley_terrestrial_2019; @li_mapping_2019]. In contrast, some studies suggest that the highest values of net primary productivity ($NPP$) may be found in temperate forests [@luyssaert_co_2007; @huston_global_2009], while others find $NPP$ highest in the tropics and decreasing with latitude [@simova_enigma_2017]. Other studies have chosen to focus exclusively on above-ground net primary productivity ($ANPP$), finding evidence of a weak negative relationship between $ANPP$ and latitude [@huston_global_2009; @gillman_latitude_2015].

The latitudinal gradient in FACF is primarily driven by climate, which is a significant driver of FACF across broad spatial scales [@luyssaert_co_2007; @cleveland_relationships_2011; @hursh_sensitivity_2017]. The majority of studies have focused on exploring the relationships of FACF to mean annual temperature (MAT) and mean annual precipitation (MAP), as the most commonly reported site-level climate variables. There is strong evidence that both MAT and MAP show significant positive relationships with FACF [@chu_does_2016]. However, as with latitude, the shape of those relationships is not always clear, and, again, is complicated by the use of different measures of FACF across studies. Various measures of primary productivity {**FACF?**} saturate at high levels of MAP, though the saturation points identified vary from 1500mm [@luyssaert_co_2007] up to 2445mm MAP [@schuur_productivity_2003]. Studies of the influence of MAT on productivity {**FACF?**} are less conclusive. Luyssaert et al. [-@luyssaert_co_2007] examined GPP and NPP and found that, while $GPP$ increases linearly with MAT, NPP saturates at around 10$^\circ$C MAT. In contrast, Larjavaara and Muller-Landau [-@larjavaara_temperature_2012], find that increases in $GPP$ saturate at approximately 25$^\circ$C MAT, while Schuur [-@schuur_productivity_2003] shows that $NPP$ increases linearly with temperature. Taylor et al. [-@taylor_temperature_2017] showed a positive interaction between MAT and MAP in shaping tropical forest productivity, such that high rainfall had a negative effect on productivity in cooler climates, compared to a positive effect in warmer climates. Such complicated dynamics play out with belowground fluxes as well, with typically dominant soil temperature mediated by moisture and carbon supply (Hursh et al. 2017, http://dx.doi.org/10.1111/gcb.13489; Xu et al. 2016 http://dx.doi.org/10.1016/j.jplph.2016.08.007).

MAT and MAP are very coarse measures of climate, and so fail to capture much variation in climate on an intra-annual scale, including the effects of factors such as growing season length, number of frost-free days, temperature seasonality, and dry season length. Some studies have suggested that the apparently strong relationship between MAT and FACFs is actually a consequence of the correlation between MAT and growing season length [@kerkhoff_plant_2005; @malhi_productivity_2012; @michaletz_convergence_2014; @michaletz_drivers_2018]. Kerkhoff et al. [-@kerkhoff_plant_2005] and Michaletz et al. [-@michaletz_convergence_2014] find that, within the growing season, there is no significant relationship between primary productivity and MAT, suggesting that the effect of temperature is due to increased length of growing season, rather than an inherent influence of temperature on FACF.

In addition, FACF can be influenced by other, often interactive, factors across a range of scales [@cleveland_relationships_2011]. On a local scale, stand age [@litton_carbon_2007; @gillman_latitude_2015], biodiversity [@liang_positive_2016], management [@simova_enigma_2017], nutrient availability [@aragao_above_2009], and altitude [@girardin_net_2010; @malhi_variation_2017] all impact FACF. There is evidence that FACFs also respond to variables such as cloud cover [@taylor_temperature_2017], solar radiation [@fyllas_solar_2017], and potential evapotranspiration [@kerkhoff_plant_2005].

@@

\renewcommand{\arraystretch}{2}
```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
hypothesis_table <- read.csv("hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
hypothesis_table[1,1] <- paste0(hypothesis_table[1,1], footnote_marker_number(1),"$^{,}$",footnote_marker_number(2),"$^{,}$", footnote_marker_number(3),"$^{,}$", footnote_marker_number(10))
hypothesis_table[2,1] <- paste0(hypothesis_table[2,1], footnote_marker_number(1),"$^{,}$",footnote_marker_number(4),"$^{,}$",footnote_marker_number(9))
hypothesis_table[3,1] <- paste0(hypothesis_table[3,1], footnote_marker_number(1),"$^{,}$",footnote_marker_number(4))
hypothesis_table[4,1] <- paste0(hypothesis_table[4,1], footnote_marker_number(5))
hypothesis_table[10,1] <- paste0(hypothesis_table[10,1], footnote_marker_number(6),"$^{,}$",footnote_marker_number(7),"$^{,}$", footnote_marker_number(8))
hypothesis_table[11,1] <- paste0(hypothesis_table[11,1], footnote_marker_number(7),"$^{,}$", footnote_marker_number(8))
hypothesis_table[12,1] <- paste0(hypothesis_table[12,1], footnote_marker_number(8))
kable(hypothesis_table, format = "latex", booktabs = TRUE, caption = "Summary of research questions, corresponding hypotheses, and results. Statistically signficant support for/ rejection of hypotheses is indicated with 'yes'/'no', parentheses indicate partial overall support/rejection of hypotheses across all FACF, and '-' indicates no significant relationship.", escape = F) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12) %>%
  pack_rows("Q1. How do FACF vary with latitude?", 1, 1, latex_gap_space = "1em", colnum = 5, hline_before = FALSE) %>%
  pack_rows("Q2. How do FACF vary with MAT and MAP?", 2, 4, latex_gap_space = "1em", colnum = 5, hline_before = TRUE) %>%
  pack_rows("Q3. How are FACF related to other climate variables?", 5, 7, latex_gap_space = "1em", colnum = 5, hline_before = TRUE) %>%
  pack_rows("Q4. How does seasonality influence FACF?", 8, 11, latex_gap_space = "1em", colnum = 5, hline_before = TRUE) %>%
  pack_rows("Q5. When standardised by growing season length, how do FACF vary with climate?", 12, 15, latex_gap_space = "1em", colnum = 5, hline_before = TRUE) %>%
  add_header_above(c(" ",  " ", "Forest autotrophic carbon fluxes (FACF)" = 9, " ")) %>%
  #column_spec(2:11, width = "1.8cm") %>%
  kableExtra::landscape() %>%
  kableExtra::footnote(number = c("Luyssaert et al. (2007)", "Gillman et al. (2015)", "Simova and Storch (2017)", "Schuur (2003)", "Taylor et al. (2016)", "Malhi (2012)", "Michaletz et al. (2014)", "Chu et al. (2016)", "Piao et al. (2010)", "Huston & Wolverton (2009)"), footnote_as_chunk = TRUE)
```

>>>>

Carbon cycling in forests worldwide provides the energetic basis for sustaining the majority of Earth's terrestrial biodiversity and many human populations (**[MEA 2005](http://www.millenniumassessment.org/documents/document.354.aspx.pdf)**; **REFS**), while strongly influencing atmospheric CO~2~ and climate [@bonan_forests_2008]. Forests' autotrophic carbon fluxes (FACF)--that is, carbon fixation, allocation, and metabolism by trees and other primary producers--sets the energy ultimately available to heterotrophic organisms (including microbes), in turn influencing their abundance (e.g., **[Zak et al. 1994](http://esajournals.onlinelibrary.wiley.com/doi/abs/10.2307/1940888)**; [**Niedziałkowska et al. 2010](https://doi.org/10.2980/17-1-3310)**) and possibly diversity [@waide_relationship_1999; **Chu et al. 2019 (DOI: 10.1111/ele.13175)**]. FACF influences all organic matter stocks in forest ecosystems and is linked to cycling of energy, water, and nutrients.

@@

FACF decrease with latitude, but it remains unclear whether and how the shape of this relationship varies between fluxes. FACF are lowest in the boreal regions, and increase into the temperate regions [@luyssaert_co_2007; @huston_global_2009; @beer_terrestrial_2010; @piao_forest_2010; @jung_global_2011]. However, evidence is inconclusive on whether primary productivity continues to increase into the tropics. Evidence for this is further complicated by the fact that different studies use different measures of productivity to explore these relationships. For example, modelling of global terrestrial ecosystem gross primary productivity ($GPP$) through upscaling and calibration of eddy flux measurements indicates that GPP peaks in tropical forests [@beer_terrestrial_2010; @jung_global_2011; @badgley_terrestrial_2019; @li_mapping_2019]. In contrast, some studies suggest that the highest values of net primary productivity ($NPP$) may be found in temperate forests [@luyssaert_co_2007; @huston_global_2009], while others find $NPP$ highest in the tropics and decreasing with latitude [@simova_enigma_2017]. Other studies have found that aboveground net primary productivity ($ANPP$) decreases weakly with latitude [@huston_global_2009; @gillman_latitude_2015].

The latitudinal gradient in FACF, along with altitudinal gradients [@girardin_net_2010; @malhi_variation_2017], is primarily driven by climate, which is a significant driver of FACF across broad spatial scales [@luyssaert_co_2007; @cleveland_relationships_2011; @hursh_sensitivity_2017]. The majority of studies have focused on exploring the relationships of FACF to mean annual temperature (MAT) and mean annual precipitation (MAP), as the most commonly reported site-level climate variables. There is strong evidence that both MAT and MAP show significant positive relationships with FACF [@chu_does_2016]. However, as with latitude, the shape of those relationships is not always clear, and, again, is complicated by the use of different measures of FACF across studies. Various measures of primary productivity {**FACF?**} saturate at high levels of MAP, though the saturation points identified vary from 1500mm [@luyssaert_co_2007] up to 2445mm MAP [@schuur_productivity_2003]. Studies of the influence of MAT on productivity {**FACF?**} are less conclusive. Luyssaert et al. [-@luyssaert_co_2007] examined GPP and NPP and found that, while $GPP$ increases linearly with MAT, NPP saturates at around 10$^\circ$C MAT. In contrast, Larjavaara and Muller-Landau [-@larjavaara_temperature_2012], find that increases in $GPP$ saturate at approximately 25$^\circ$C MAT, while Schuur [-@schuur_productivity_2003] finds that $NPP$ increases linearly with temperature. Taylor et al. [-@taylor_temperature_2017] showed a positive interaction between MAT and MAP in shaping tropical forest productivity, such that high rainfall had a negative effect on productivity in cooler climates, compared to a positive effect in warmer climates. Such complicated dynamics play out with belowground fluxes as well, with typically dominant soil temperature mediated by moisture and carbon supply (**Hursh et al. 2017, http://dx.doi.org/10.1111/gcb.13489; Xu et al. 2016 http://dx.doi.org/10.1016/j.jplph.2016.08.007**).

MAT and MAP are very coarse measures of climate, and so fail to capture much variation in climate on an intra-annual scale, including the effects of factors such as growing season length, number of frost-free days, temperature seasonality, and dry season length. Some studies have suggested that the apparently strong relationship between MAT and FACFs is actually a consequence of the correlation between MAT and growing season length [@kerkhoff_plant_2005; @malhi_productivity_2012; @michaletz_convergence_2014; @michaletz_drivers_2018]. Kerkhoff et al. [-@kerkhoff_plant_2005] and Michaletz et al. [-@michaletz_convergence_2014] find that, within the growing season, there is no significant relationship between net primary productivity (loosely defined) and MAT, suggesting that the influence of temperature may be limited to determining the length of the frost-free growing season.

**the following doesn't fit very well...**  
In addition, FACF can be influenced by other, often interactive, factors across a range of scales [@cleveland_relationships_2011]. On a local scale, stand age [@litton_carbon_2007; @gillman_latitude_2015], biodiversity [@liang_positive_2016], management [@simova_enigma_2017], and nutrient availability [@aragao_above_2009]. There is evidence that FACFs also respond to variables such as cloud cover [@taylor_temperature_2017], solar radiation [@fyllas_solar_2017], and potential evapotranspiration [@kerkhoff_plant_2005].

@@

\renewcommand{\arraystretch}{2}
```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
hypothesis_table <- read.csv("hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
hypothesis_table[1,1] <- paste0(hypothesis_table[1,1], footnote_marker_number(1),"$^{,}$",footnote_marker_number(2),"$^{,}$", footnote_marker_number(3),"$^{,}$", footnote_marker_number(10))
hypothesis_table[2,1] <- paste0(hypothesis_table[2,1], footnote_marker_number(1),"$^{,}$",footnote_marker_number(4),"$^{,}$",footnote_marker_number(9))
hypothesis_table[3,1] <- paste0(hypothesis_table[3,1], footnote_marker_number(1),"$^{,}$",footnote_marker_number(4))
hypothesis_table[4,1] <- paste0(hypothesis_table[4,1], footnote_marker_number(5))
hypothesis_table[10,1] <- paste0(hypothesis_table[10,1], footnote_marker_number(6),"$^{,}$",footnote_marker_number(7),"$^{,}$", footnote_marker_number(8))
hypothesis_table[11,1] <- paste0(hypothesis_table[11,1], footnote_marker_number(7),"$^{,}$", footnote_marker_number(8))
hypothesis_table[12,1] <- paste0(hypothesis_table[12,1], footnote_marker_number(8))
kable(hypothesis_table, format = "latex", booktabs = TRUE, caption = "Summary of research questions, corresponding hypotheses, and results. Statistically signficant support for/ rejection of hypotheses is indicated with 'yes'/'no', and '-' indicates no significant relationship. Parentheses indicate partial overall support or rejection of hypotheses across all fluxes considered.", escape = F) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12) %>%
  pack_rows("Q1. How do FACF vary with latitude?", 1, 1, latex_gap_space = "1em", colnum = 5, hline_before = FALSE) %>%
  pack_rows("Q2. How do FACF vary with MAT and MAP?", 2, 4, latex_gap_space = "1em", colnum = 5, hline_before = TRUE) %>%
  pack_rows("Q3. How are FACF related to other climate variables?", 5, 7, latex_gap_space = "1em", colnum = 5, hline_before = TRUE) %>%
  pack_rows("Q4. How does seasonality influence FACF?", 8, 11, latex_gap_space = "1em", colnum = 5, hline_before = TRUE) %>%
  pack_rows("Q5. When standardised by growing season length, how do FACF vary with climate?", 12, 15, latex_gap_space = "1em", colnum = 5, hline_before = TRUE) %>%
  add_header_above(c(" ",  " ", "Forest autotrophic carbon fluxes (FACF)" = 9, " ")) %>%
  #column_spec(2:11, width = "1.8cm") %>%
  kableExtra::landscape() %>%
  kableExtra::footnote(number = c("Luyssaert et al. (2007)", "Gillman et al. (2015)", "Simova and Storch (2017)", "Schuur (2003)", "Taylor et al. (2016)", "Malhi (2012)", "Michaletz et al. (2014)", "Chu et al. (2016)", "Piao et al. (2010)", "Huston & Wolverton (2009)"), footnote_as_chunk = TRUE)
```
